\documentclass[11pt]{letter}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}

\signature{Evan Paul}
\address{evan@evanpaul.us \\ Nicasio, CA}

\begin{document}

\begin{letter}{%
Editorial Office\\
Physical Review Letters\\
American Physical Society}

\opening{Dear Editors,}

I am submitting the enclosed manuscript, ``Strange Attractors in Gradient Descent: Data Structure and Loss Geometry Control Fractal Dimension,'' for consideration as a Letter in Physical Review Letters.

\textbf{Summary of the result.} We measure the fractal dimension of neural network training trajectories in function space using the Grassberger-Procaccia correlation dimension $D_2$. A convolutional network trained on CIFAR-10 with full-batch gradient descent produces a strange attractor with $D_2 = 3.67 \pm 0.08$ (conservative lower bound from $N \approx 400$ points; convergence analysis at $N = 3200$ yields $D_2 \approx 5$). Cross-architecture controls (five conditions, 10 seeds each) reveal that structured data is necessary for multi-dimensional chaos; architecture modulates only the learning-rate threshold. A continuous label-noise sweep reveals two distinct mechanisms controlling attractor dimension: at moderate learning rates, structured labels create oscillatory modes and degrading them smoothly reduces $D_2$; near the Edge of Stability, loss-surface roughness from random labels sustains high-dimensional dynamics independently. Peak perturbation sensitivity and peak geometric complexity occur at different learning rates across all architectures, with the richest structure in the transition zone between chaotic and convergent regimes.

\textbf{Significance and broad appeal.} This work connects the empirical phenomenology of neural network training (Edge of Stability, sharpness oscillations, period-doubling) to concepts from classical dynamical systems theory. The result that training trajectories are strange attractors whose dimension is shaped by two distinct mechanisms---data structure and loss-surface geometry---reframes optimization as a problem in nonlinear dynamics, and provides a geometric language (fractal dimension, attractor structure, dynamical transitions) for describing what happens during learning. The finding that the sign of $D_2$'s response to label noise reverses between dynamical regimes is, to our knowledge, a new observation. We believe this is of interest to physicists working in nonlinear dynamics, statistical mechanics of learning, and complex systems, beyond the immediate machine learning community.

\textbf{Measurement validation.} The $D_2$ pipeline is calibrated against 11 known dynamical systems (tori, H\'{e}non, Lorenz, R\"{o}ssler, Mackey-Glass) at the production trajectory length. A $D_2(N)$ convergence analysis confirms reported values are conservative lower bounds. Persistent homology independently confirms fractal topology for all architectures. The supplemental material provides convergence audits for all measurement parameters ($\varepsilon$, number of held-out inputs, trajectory length) and data for the full label-noise sweep.

\textbf{Disclosure of AI tool usage.} In accordance with APS policy, I wish to provide a full and transparent accounting of AI involvement in this work.

AI language models (Claude, Anthropic) were used as interactive research tools throughout the project. This went beyond manuscript polishing and included:

\begin{itemize}
\item \textit{Experimental methodology:} The function-space Lyapunov protocol, the perturbation sensitivity analysis that identified the $\varepsilon = 10^{-8}$ numerical artifact, the cross-experiment design isolating data complexity from architecture, and the $D_2$ calibration pipeline against known attractors were developed through iterative dialogue with AI.
\item \textit{Code development:} Experiment scripts (training loops, Grassberger-Procaccia implementation, persistent homology analysis, figure generation) were co-developed with AI assistance.
\item \textit{Theoretical connections:} AI helped identify the mapping between the observed dynamics and classical routes to chaos (Ruelle-Takens, related phenomenology), and helped locate relevant literature (Pittorino et al., Storm et al.).
\item \textit{Manuscript preparation:} Multiple drafts were produced with AI assistance, including structural organization, prose revision, and \LaTeX{} formatting.
\end{itemize}

What was mine throughout: the original research question (whether training dynamics have toroidal/fractal geometry), all decisions about which experiments to run and which directions to pursue, the choice to measure correlation dimension rather than other invariants, the interpretation of results, and the decision to target PRL. I ran all experiments, verified all data, and take full responsibility for every claim in the manuscript.

I recognize that this level of AI involvement exceeds what APS policy currently addresses under ``light editing.'' I have chosen full transparency because I believe the scientific community benefits from honest disclosure of evolving research practices, and because the work stands on its reproducible experimental results regardless of how the methodology was developed. All code and data are publicly available at \url{https://github.com/gnomevan/attractor}.

\textbf{Referee suggestions.}
\begin{itemize}
\item Jeremy Cohen (Carnegie Mellon / Google DeepMind) --- first author of the Edge of Stability work on which our learning-rate parameterization is built.
\item Bernhard Mehlig (University of Gothenburg) --- senior author of Storm et al.\ on finite-time Lyapunov exponents in neural networks; published in PRL.
\item Riccardo Zecchina (Bocconi University) --- senior author of Pittorino et al.\ on loss landscape topology; broad expertise in statistical physics of learning.
\end{itemize}

\closing{Sincerely,}

\end{letter}
\end{document}
